---
title: Problem set 9
date: 2025-04-13
---

This problem set explores the analysis of high dimensional data in the application area of spatial transcriptomics. For reference, consult the following papers:

* [Robust decomposition of cell type mixtures in spatial transcriptomics](https://www.nature.com/articles/s41587-021-00830-w)
* [Cell type-specific inference of differential expression in spatial transcriptomics](https://www.nature.com/articles/s41592-022-01575-3)

### Load in the data

We begin by downloading the data. Hint: run this once and leave it as `eval = FALSE` in your script.

```{r, eval = FALSE}
# download data
dir.create('data')
file_list <- c('gene_counts', 'cell_coords', 'neuron_cells', 'other_cells',
               'positive_genes', 'negative_genes', 'other_genes')
for(filename in file_list) {
  download.file(paste0('https://github.com/dmcable/BIOSTAT620/raw/main/data/p9/',filename,'.rds'), 
                destfile = paste0('data/',filename,'.rds'))
}
```
Next, we load in the data and packages (note that `counts` is a sparse matrix):

```{r}
# required packages
library(ggplot2)
library(Matrix)
library(tidyverse)

# load data
counts <- readRDS('data/gene_counts.rds') # gene counts matrix
coords <- readRDS('data/cell_coords.rds') # cell coords matrix

neuron_cells <- readRDS('data/neuron_cells.rds') # list of neuron cell barcodes.
other_cells <- readRDS('data/other_cells.rds') # list of non-neuron cell barcodes.

positive_genes <- readRDS('data/positive_genes.rds') # list of genes specific for neurons
negative_genes <- readRDS('data/negative_genes.rds') # list of genes specific for not neurons
other_genes <- readRDS('data/other_genes.rds') # selected list of other genes

all_genes <- c(positive_genes, negative_genes, other_genes) # all genes
small_counts <- as.matrix(counts[all_genes, c(neuron_cells, other_cells)]) # subset of the counts
```

### Data visualization

1. Begin by visualizing the coordinates of the cells in 2D space. Hint: this information is contained in the `coords` dataframe.

```{r}
set.seed(2025)
plot(coords,
     pch = 16, 
     col = "blue")
```

2. Let's now visualize the expression of the Hpca gene, which is specific for neurons. Begin by creating a histogram of the counts of the gene. Hint: examine the `counts` matrix.

```{r}
"Hpca" %in% positive_genes

grep("hpca", positive_genes, ignore.case = TRUE, value = TRUE)

hpca_expr <- small_counts["Hpca", ]

hpca_expr <- as.vector(hpca_expr)

hist(hpca_expr,
     breaks = 50,
     main = "Expression of Hpca Gene (Neuronal Marker)",
     xlab = "Hpca Expression Count",
     col = "steelblue",
     border = "white")

hist(log1p(hpca_expr),
     breaks = 50,
     main = "Log Expression of Hpca Gene (Neuronal Marker)",
     xlab = "log(1 + Hpca Expression)",
     col = "darkorange",
     border = "white")
```

3. Make a spatial plot of the gene expression of this gene. Use the cell coordinates as the spatial coordinates, and use gene expression counts as the color. 4

```{r}

max_expr <- 20

selected_cells <- c(neuron_cells, other_cells)

coords_sub <- coords[selected_cells, ]

coords_df <- as.data.frame(coords_sub)
colnames(coords_df) <- c("x", "y")

hpca_expr <- as.vector(small_counts["Hpca", ])

max_expr <- quantile(hpca_expr, 0.99)
hpca_expr_capped <- pmin(hpca_expr, max_expr)

coords_df$hpca_expr <- hpca_expr_capped


ggplot(coords_df, aes(x = x, y = y, color = hpca_expr_capped)) +
  geom_point(size = 0.7) +
  scale_color_viridis_c(option = "plasma", name = "Hpca Expr", limits = c(0, max_expr)) +
  labs(title = "Spatial Plot of Hpca Gene Expression",
       x = "X Coordinate", y = "Y Coordinate") +
  theme_minimal()
```
### Distance

4. Now, make a spatial plot of the cell types. Create a dataframe `plot_df` containing only the neurons and the non-neurons. Plot the neurons and non-neurons as two different colors. Hint: the neuron barcodes are in `neuron_cells`, and the non-neuron barcodes are in `other_cells`. 

```{r}
selected_cells <- c(neuron_cells, other_cells)

coords_sub <- coords[selected_cells, ]

plot_df <- as.data.frame(coords_sub)
colnames(plot_df) <- c("x", "y")

plot_df$barcode <- rownames(plot_df)

plot_df$cell_type <- ifelse(plot_df$barcode %in% neuron_cells, "Neuron", "Non-neuron")

ggplot(plot_df, aes(x = x, y = y, color = cell_type)) +
  geom_point(size = 0.7, alpha = 0.8) +
  scale_color_manual(values = c("Neuron" = "blue", "Non-neuron" = "yellow")) +
  labs(title = "Spatial Plot of Cell Types",
       x = "X Coordinate", y = "Y Coordinate",
       color = "Cell Type") +
  theme_minimal()
```

5. Using only the cells selected above in `plot_df`, compute the distance matrix pairwise between each cell (hint: use `rdist` from `fields`). and compute the k-nearest-neighbors with `K = 25` (hint use `get.knn` from `FNN`). Choose the first cell, and visualize the spatial locations of the k nearest neighbors relative to that cell.

```{r}
library(fields)
library(FNN)   
coords_mat <- as.matrix(plot_df[, c("x", "y")]) 
dist_mat <- rdist(coords_mat)

knn_result <- get.knn(coords_mat, k = 25)

first_index <- 1

neighbor_indices <- knn_result$nn.index[first_index, ]

first_cell <- plot_df[first_index, ]
neighbors <- plot_df[neighbor_indices, ]

ggplot() +
  geom_point(data = plot_df, aes(x = x, y = y), color = "grey", size = 0.5) + 
  geom_point(data = neighbors, aes(x = x, y = y), color = "yellow", size = 1.5) + 
  geom_point(data = first_cell, aes(x = x, y = y), color = "blue", size = 2) + 
  labs(title = "25 Nearest Neighbors of the First Cell",
       x = "X Coordinate", y = "Y Coordinate") +
  theme_minimal()
```

6. For each cell, calculate the proportion of neighbors that are neurons, and visualize these proportions spatially. 

```{r}
is_neuron <- ifelse(plot_df$cell_type == "Neuron", 1, 0)

neighbor_idx <- knn_result$nn.index

neuron_proportions <- apply(neighbor_idx, 1, function(idx) {
  mean(is_neuron[idx])
})

plot_df$neuron_neighbor_proportion <- neuron_proportions

ggplot(plot_df, aes(x = x, y = y, color = neuron_neighbor_proportion)) +
  geom_point(size = 0.7) +
  scale_color_viridis_c(name = "Prop. Neuron\nin Neighbors", limits = c(0, 1)) +
  labs(title = "Proportion of Neuronal Neighbors per Cell",
       x = "X Coordinate", y = "Y Coordinate") +
  theme_minimal()
```

### Smoothing

7. Using `loess`, fit a 2D spatial smoothing function to the neuron proportion values. Use `degree = 1` and `span = 0.05`. Create a spatial plot with the color as the smooth fit. 

```{r}
loess_fit <- loess(neuron_neighbor_proportion ~ x + y,
                   data = plot_df,
                   degree = 1,
                   span = 0.05,
                   control = loess.control(surface = "direct"))

plot_df$loess_smooth <- predict(loess_fit, newdata = plot_df)

ggplot(plot_df, aes(x = x, y = y, color = loess_smooth)) +
  geom_point(size = 0.7) +
  scale_color_viridis_c(name = "Smoothed\nNeuron Prop", limits = c(0, 1)) +
  labs(title = "Smoothed 2D LOESS Fit of Neuron Proportions",
       x = "X Coordinate", y = "Y Coordinate") +
  theme_minimal()
```

8. Visualize a 1-dimensional section of the smooth function. Filter for `y` within `50` of `3000`. Plot both the proportion and the fitted smooth values in two separate colors.

```{r}
section_df <- plot_df %>% 
  filter(abs(y - 3000) <= 50)

ggplot(section_df, aes(x = x)) +
  geom_point(aes(y = neuron_neighbor_proportion), color = "orange", size = 1) +
  geom_point(aes(y = loess_smooth), color = "blue", size = 1) +
  labs(title = "1D Section at y ≈ 3000",
       x = "X Coordinate",
       y = "Neuron Proportion / LOESS Fit") +
  theme_minimal()
```

### Dimension reduction

9. Using a subset of the counts (`small_counts`), perform a PCA (hint: use `prcomp`). Plot the first principal component (hint: check the `rotation` variable of the `prcomp` object) vs the total cellular count.  

```{r}
cell_totals <- colSums(small_counts)

pca_result <- prcomp(t(small_counts), center = TRUE, scale. = TRUE)

pc1 <- pca_result$x[, 1]

plot(cell_totals, pc1,pch = 16, col = "blue")
```
Comment on your finding:

This plot shows a strong negative linear relationship between: PC1 and total cellular count

10. To mitigate this issue, we will normalize `small_counts` by dividing each cell by the sum of the total counts. Hint: use the `sweep` function. Repeate the PCA with the normalized count matrix. Make a plot of percent of variance explained for each of the first 500 PCs (threshold this plot at 5% variance explained). 

```{r}
cell_totals <- colSums(small_counts)
norm_counts <- sweep(small_counts, 2, cell_totals, FUN = "/")

pca_norm <- prcomp(t(norm_counts), center = TRUE, scale. = TRUE)

var_explained <- (pca_norm$sdev)^2 / sum(pca_norm$sdev^2)

var_explained_500 <- var_explained[1:500]
var_explained_500 <- pmin(var_explained_500, 0.05)

plot(var_explained_500 * 100,
     type = "h",
     xlab = "Principal Component",
     ylab = "Variance Explained (%)",
     main = "Variance Explained by First 500 PCs (Capped at 5%)")
```

11. Make spatial plots visualizing the first four PCs in spatial coordinates.

```{r}
pca_scores <- pca_norm$x[, 1:4]

plot_df$PC1 <- pca_scores[, 1]
plot_df$PC2 <- pca_scores[, 2]
plot_df$PC3 <- pca_scores[, 3]
plot_df$PC4 <- pca_scores[, 4]

ggplot(plot_df, aes(x = x, y = y, color = PC1)) +
  geom_point(size = 0.7) +
  scale_color_viridis_c() +
  labs(title = "Spatial Plot of PC1") +
  theme_minimal()

ggplot(plot_df, aes(x = x, y = y, color = PC2)) +
  geom_point(size = 0.7) +
  scale_color_viridis_c() +
  labs(title = "Spatial Plot of PC2") +
  theme_minimal()

ggplot(plot_df, aes(x = x, y = y, color = PC3)) +
  geom_point(size = 0.7) +
  scale_color_viridis_c() +
  labs(title = "Spatial Plot of PC3") +
  theme_minimal()

ggplot(plot_df, aes(x = x, y = y, color = PC4)) +
  geom_point(size = 0.7) +
  scale_color_viridis_c() +
  labs(title = "Spatial Plot of PC4") +
  theme_minimal()
```

12. For the first 20 pcs, compute the average weight for each of the following sets of genes: `positive_genes`, `negative_genes`, and `other_genes`. Create a `geom_point` plot of these with PC index as the x-axis, average weight as the y axis, and gene list type as the color.

```{r}
rotation_mat <- pca_norm$rotation[, 1:20]

pos_avg <- colMeans(rotation_mat[positive_genes, ])
neg_avg <- colMeans(rotation_mat[negative_genes, ])
oth_avg <- colMeans(rotation_mat[other_genes, ])

df_plot <- tibble(
  PC = rep(1:20, times = 3),
  avg_weight = c(pos_avg, neg_avg, oth_avg),
  gene_set = rep(c("Positive", "Negative", "Other"), each = 20)
)

ggplot(df_plot, aes(x = PC, y = avg_weight, color = gene_set)) +
  geom_point() +
  labs(title = "Average Gene Loadings on PCs by Gene Set",
       x = "Principal Component",
       y = "Average Weight") +
  theme_minimal()
```

13. Now, remake the previous plot for the first five PCs and include standard error bars.

```{r}
rotation_mat <- pca_norm$rotation[, 1:5]

pos_df <- as_tibble(rotation_mat[positive_genes, ]) %>%
  pivot_longer(cols = everything(), names_to = "PC", values_to = "weight") %>%
  group_by(PC) %>%
  summarise(mean = mean(weight), se = sd(weight) / sqrt(n()), .groups = "drop") %>%
  mutate(gene_set = "Positive")

neg_df <- as_tibble(rotation_mat[negative_genes, ]) %>%
  pivot_longer(cols = everything(), names_to = "PC", values_to = "weight") %>%
  group_by(PC) %>%
  summarise(mean = mean(weight), se = sd(weight) / sqrt(n()), .groups = "drop") %>%
  mutate(gene_set = "Negative")

oth_df <- as_tibble(rotation_mat[other_genes, ]) %>%
  pivot_longer(cols = everything(), names_to = "PC", values_to = "weight") %>%
  group_by(PC) %>%
  summarise(mean = mean(weight), se = sd(weight) / sqrt(n()), .groups = "drop") %>%
  mutate(gene_set = "Other")

df_plot <- bind_rows(pos_df, neg_df, oth_df)

ggplot(df_plot, aes(x = as.numeric(str_remove(PC, "PC")), y = mean, color = gene_set)) +
  geom_point() +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se), width = 0.2) +
  labs(title = "Average PC Weights with SE for First 5 PCs",
       x = "Principal Component",
       y = "Average Weight") +
  theme_minimal()
```


Which of the first 5 PCs appear to have statistically significant differences across gene types?

I would say PC1 appears to have statistically significant differences across gene types. Because we can see that PCs 2 to 5 do not show clear separation based on what we draw on the graph.
